---
title: "research_question_1"
format: pdf
editor: visual
---

```{r load-packages, message = FALSE, warning = FALSE}
library(tidyverse)
library(readxl)
library(dplyr)
library(ggplot2)
```

```{r data-1, warning=FALSE}
#Read the dataset, select the columns to be used and drop the missing data
energy_rq1 <- read_excel("owid-energy-data.xlsx") %>%
  select(country, year, gdp, population, carbon_intensity_elec,fossil_share_elec,electricity_generation) %>%
  mutate(gdp_per_capita = gdp / population) %>%
  filter(!is.na(carbon_intensity_elec),
         !is.na(gdp_per_capita))

glimpse(energy_rq1)
```

**Research question 1:**

**Does a country’s GDP per capita significantly affect the carbon intensity of its electricity generation, and does this relationship differ between high-income and low-income countries?**

-   Outcome variable (include the name/description and type of variable):
    -   `carbon_intensity_elec` **(continuous)**：measures the amount of carbon dioxide emitted per unit of electricity generated (grams of CO₂ per kilowatt-hour). It captures how carbon-efficient a country’s electricity production is — lower values indicate cleaner, more renewable-based power systems.
